Accelerated and interpretable oblique random survival forests

### 2022-11-23 paper/arxiv/main.tex
working on jcgs submission

This Section introduces negation VI, which is similar to permutation VI in that it measures how much a model’s prediction error increases when a variable’s role in the model is de-stabilized. Specifically, negation VI measures the increase in an oblique RF's prediction error after flipping the sign of all coefficients linked to a variable (\ie, negating them). As the magnitude of a coefficient increases, so does the probability that negating it will change the oblique RF's predictions. For the current study, we use Harrell's concordance (C)-statistic \citep{harrell1982evaluating} to measure change in prediction error when computing negation VI.

Negation VI has several helpful characteristics. First, negation VI generalizes to any oblique RF (\ie, not just RSFs) using any valid error function, making it both general and flexible.\footnote{The \texttt{aorsf} package enables customized functions to be applied in lieu of the default C-statistic (see \texttt{?aorsf::orsf\_vi\_negate})} Second, since the coefficients in each non-leaf node of an oblique tree are adjusted for the accompanying predictors, negation VI may provide better estimation of VI in the presence of correlated variables compared to standard VI techniques. Third, unlike permutation, negation is non-random and hence reproducible without setting a random seed. Additionally, since negation VI does not permute variables, the analyst need not worry about impossible combinations of predictors that may occur when one predictor is randomly permuted, such as having a negative status for type 2 diabetes and having Hemoglobin A1c level $\geq$ 6.5\% (a value indicative of type 2 diabetes) as a result of randomly permuting the values of Hemoglobin A1c.

@@

Sections \ref{sec:bm_pred} and \ref{sec:bm_vi} present numerical experiments examining the accelerated oblique RSF and negation VI, respectively. The code used to run these experiments is available online at \href{https://github.com/bcjaeger/aorsf-bench}{https://github.com/bcjaeger/aorsf-bench}. All analyses were conducted using R version 4.1.3 and coordinated by the \texttt{targets} R package \citep{targets}. To standardize comparisons of computational efficiency, all learners and VI techniques used up to 4 processing units.

\subsection{Benchmark of prediction accuracy and computational efficiency} \label{sec:bm_pred}

The aim of this numeric experiment is to evaluate and compare the accelerated oblique RSF with its predecessor (the oblique RSF from the \texttt{obliqueRSF} R package) and with other machine learning algorithms for risk prediction. Inferences drawn from this experiment include equivalence and inferiority tests based on Bayesian linear mixed models.

@@

We used a collection of 21 data sets containing a total of 35 risk prediction tasks (tasks per data set ranged from one to four) to benchmark the accelerated oblique RSF versus other learners. Participant-level data from the GUIDE-IT and SPRINT clinical trials and the ARIC, MESA, and JHS community cohort studies was obtained from the National Institute of Health Biologic Specimen and Data Repository Coordinating Center (BioLINCC). Designs and protocols for these studies have been made available \citep{aric1989atherosclerosis, bild2002multi, felker2017effect, sprint2015randomized, taylor2005toward}. All other datasets were publicly available and obtained through R packages (see Appendix A.1). Across all prediction tasks, the number of observations ranged from 137 to 17,549 (median: 1,384), the number of predictors ranged from 7 to 1,692 (median: 41), and the percentage of censored observations ranged from 5.26 to 97.7 (median: 78.1) (\tabrefAppendix{tab:datasets}).

@@

For each risk prediction task, we completed 25 runs of Monte-Carlo cross validation. In each run, we used a random sample containing 50\% of the available data for training and the remaining 50\% for testing each of the learners described in \secref{sec:learners}. Then, for each learner, we computed the IPA, time-dependent C-statistic, and computational time required to fit a prediction model and compute risk predictions. If any learner failed to obtain predictions on any particular split of data\footnote{For example, when the prediction task was to predict risk of death in the ACTG 320 clinical trial (26 events total), some splits did not leave enough events in the training data to fit complex learners such as the neural network}, the results for that split were omitted from downstream analyses.

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest IPA in 20 out of 35 risk prediction tasks, with an overall mean IPA of 12.6 (Figure \ref{fig:bm_pred_viz_ibs}). Compared to the learner with the second highest mean IPA (\texttt{cif-standard}), \texttt{aorsf-fast}'s mean was 1.35 points higher, a relative increase of 12.0\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected IPA was 0.99, and the posterior probability of \texttt{aorsf-fast} having a superior IPA to other learners ranged from 0.86 (versus \texttt{cif-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_ibs})

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest time-dependent C-statistic in 10 out of 35 risk prediction tasks, with an overall mean of 77.2 (Figure \ref{fig:bm_pred_viz_cstat}). Compared to the learner with the second highest mean C-statistic (\texttt{cif-standard}), \texttt{aorsf-fast}'s mean was 0.685 points higher, a relative increase of 0.896\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected time-dependent C-statistics was $>$ 0.999, and the posterior probability of \texttt{aorsf-fast} having a superior time-dependent C-statistic versus other learners ranged from 0.18 (versus \texttt{cif-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_cstat})

@@

\paragraph{Computational efficiency}

Overall, \texttt{aorsf-fast} was the second fastest learner, with an expected model development and risk prediction time about 385 milliseconds longer than \texttt{glmnet-cox} (Figure \ref{fig:bm_pred_time}). Comparing median computing times, \texttt{aorsf-fast} was 379.8 times faster than its predecessor, \texttt{obliqueRSF-net}. In addition, \texttt{aorsf-fast} was 16.1, 1.55, and 2.66 faster than axis based forests grown using the \texttt{party}, \texttt{ranger}, and \texttt{randomForestSRC} packages, respectively.

@@

Parameters that varied in the current simulation study included the number of observations (500, 1000, and 2500) and the absolute value of the maximum correlation between predictors (0.3, 0.15, and 0). Parameters that remain fixed throughout the study included the number of predictors in each class (15) and the effect size of each predictor (one standard deviation increase associated with a 64\% increase in relative risk). Using this design for simulated data, the Heller explained relative risk (95\% confidence interval) of our covariates was 88.4 (88.1, 88.7) \citep{heller2012measure} with 2,500 observations.

@@

@@

In this paper, we have developed two contributions to the oblique RSF: (1) the accelerated oblique RSF (\ie, \texttt{aorsf-fast}) and (2) negation VI. Our technique to accelerate the oblique RSF reduces the number of operations required to find linear combinations of inputs using a single iteration of Newton Raphson scoring, while our VI technique directly engages with coefficients in linear combinations of inputs to measure importance of individual variables. In numeric experiments, we found that that \texttt{aorsf-fast} is approximately 379.8 times faster than its predecessor, \texttt{obliqueRSF-net}, with a practically equivalent C-statistic. We also found that negation VI, a technique to estimate VI using the oblique RSF, detected non-linear, combination, and main effects more effectively than three standard methods to estimate VI: permutation, ANOVA, and SHAP VI. Overall, we found that estimating VI using negation instead of ANOVA increased the C-statistic for ranking a relevant variable higher than an irrelevant variable by 2.05, a relative increase of 2.78\%.

@@

\hspace{1em}aorsf-fast & 0.126 (0.110) & 0.772 (0.071) & 0.620 & 0.145\\
\hspace{1em}aorsf-cph & 0.126 (0.109) & 0.771 (0.070) & 1.912 & 0.148\\
\hspace{1em}aorsf-net & 0.124 (0.112) & 0.771 (0.071) & 78.512 & 0.143\\
\hspace{1em}obliqueRSF-net & 0.113 (0.085) & 0.771 (0.071) & 300.393 & 18.803\\
\hspace{1em}cif-standard & 0.113 (0.098) & 0.765 (0.071) & 4.293 & 6.044\\
\hspace{1em}rsf-standard & 0.113 (0.114) & 0.755 (0.075) & 2.099 & 0.199\\
\hspace{1em}cif-rotate & 0.112 (0.124) & 0.755 (0.081) & 37.695 & 8.348\\
\hspace{1em}glmnet-cox & 0.108 (0.119) & 0.757 (0.077) & 0.473 & 0.003\\
\hspace{1em}ranger-extratrees & 0.099 (0.085) & 0.762 (0.067) & 0.878 & 1.156\\
\hspace{1em}cif-extension & 0.095 (0.092) & 0.761 (0.072) & 22.321 & 6.949\\
\hspace{1em}aorsf-random & 0.090 (0.081) & 0.741 (0.063) & 1.790 & 0.141\\
\hspace{1em}xgboost-cox & 0.064 (0.100) & 0.749 (0.094) & 3.973 & 0.004\\
\hspace{1em}nn-cox & 0.047 (0.108) & 0.656 (0.138) & 15.352 & 2.060\\
\hspace{1em}xgboost-aft & --- & 0.762 (0.076) & 13.060 & 0.007\\

@@

\hspace{1em}aorsf-random & 0.028 (0.022) & 0.748 (0.038) & 0.382 & 0.034\\
\hspace{1em}ranger-extratrees & 0.028 (0.017) & 0.740 (0.036) & 0.064 & 0.146\\
\hspace{1em}obliqueRSF-net & 0.027 (0.022) & 0.746 (0.038) & 25.980 & 15.236\\
\hspace{1em}aorsf-cph & 0.025 (0.029) & 0.751 (0.042) & 0.424 & 0.033\\
\hspace{1em}cif-standard & 0.024 (0.031) & 0.744 (0.040) & 1.771 & 4.657\\
\hspace{1em}aorsf-fast & 0.024 (0.029) & 0.746 (0.043) & 0.144 & 0.032\\
\hspace{1em}cif-extension & 0.023 (0.015) & 0.722 (0.038) & 9.848 & 4.083\\
\hspace{1em}aorsf-net & 0.019 (0.034) & 0.745 (0.042) & 17.078 & 0.034\\
\hspace{1em}glmnet-cox & 0.016 (0.030) & 0.746 (0.037) & 0.179 & 0.003\\
\hspace{1em}rsf-standard & 0.005 (0.041) & 0.730 (0.042) & 0.192 & 0.061\\
\hspace{1em}cif-rotate & 0.004 (0.040) & 0.731 (0.038) & 16.095 & 3.888\\
\hspace{1em}xgboost-cox & 0.000 (0.044) & 0.751 (0.033) & 3.704 & 0.003\\
\hspace{1em}nn-cox & -0.001 (0.008) & 0.549 (0.125) & 11.075 & 0.668\\
\hspace{1em}xgboost-aft & --- & 0.737 (0.035) & 10.015 & 0.006\\

@@

\hspace{1em}obliqueRSF-net & 0.007 (0.011) & 0.823 (0.052) & 8.869 & 11.004\\
\hspace{1em}aorsf-cph & 0.007 (0.018) & 0.821 (0.060) & 0.340 & 0.018\\
\hspace{1em}aorsf-fast & 0.005 (0.019) & 0.827 (0.056) & 0.086 & 0.018\\
\hspace{1em}aorsf-random & 0.005 (0.014) & 0.788 (0.074) & 0.221 & 0.022\\
\hspace{1em}cif-extension & 0.001 (0.020) & 0.765 (0.066) & 8.436 & 3.493\\
\hspace{1em}ranger-extratrees & 0.001 (0.019) & 0.777 (0.069) & 0.044 & 0.138\\
\hspace{1em}xgboost-cox & -0.004 (0.004) & 0.500 (0.000) & 0.119 & 0.002\\
\hspace{1em}nn-cox & -0.004 (0.004) & 0.508 (0.119) & 10.867 & 0.518\\
\hspace{1em}cif-standard & -0.005 (0.025) & 0.781 (0.062) & 1.726 & 4.648\\
\hspace{1em}aorsf-net & -0.005 (0.032) & 0.806 (0.067) & 13.309 & 0.022\\
\hspace{1em}rsf-standard & -0.031 (0.051) & 0.776 (0.073) & 0.090 & 0.035\\
\hspace{1em}cif-rotate & -0.037 (0.049) & 0.707 (0.090) & 13.920 & 3.444\\
\hspace{1em}glmnet-cox & -0.065 (0.095) & 0.746 (0.098) & 0.281 & 0.002\\
\hspace{1em}xgboost-aft & --- & 0.773 (0.071) & 9.190 & 0.005\\

@@

\hspace{1em}aorsf-fast & 0.156 (0.007) & 0.810 (0.007) & 5.048 & 1.287\\
\hspace{1em}aorsf-cph & 0.153 (0.006) & 0.809 (0.007) & 14.564 & 1.300\\
\hspace{1em}aorsf-net & 0.152 (0.006) & 0.809 (0.007) & 480.098 & 1.437\\
\hspace{1em}rsf-standard & 0.150 (0.007) & 0.801 (0.007) & 9.407 & 0.991\\
\hspace{1em}obliqueRSF-net & 0.143 (0.005) & 0.811 (0.008) & 2830.398 & 356.908\\
\hspace{1em}cif-standard & 0.132 (0.005) & 0.809 (0.007) & 71.835 & 371.102\\
\hspace{1em}glmnet-cox & 0.129 (0.011) & 0.795 (0.008) & 1.473 & 0.011\\
\hspace{1em}nn-cox & 0.117 (0.029) & 0.786 (0.046) & 51.356 & 84.715\\
\hspace{1em}ranger-extratrees & 0.112 (0.005) & 0.795 (0.009) & 293.864 & 61.837\\
\hspace{1em}cif-rotate & 0.104 (0.004) & 0.783 (0.009) & 569.289 & 67.436\\
\hspace{1em}aorsf-random & 0.097 (0.005) & 0.772 (0.008) & 10.996 & 1.259\\
\hspace{1em}cif-extension & 0.069 (0.002) & 0.786 (0.009) & 169.206 & 56.091\\
\hspace{1em}xgboost-cox & 0.064 (0.017) & 0.814 (0.006) & 8.544 & 0.015\\
\hspace{1em}xgboost-aft & --- & 0.814 (0.006) & 23.781 & 0.013\\

@@

\hspace{1em}aorsf-net & 0.217 (0.006) & 0.792 (0.004) & 902.210 & 2.343\\
\hspace{1em}rsf-standard & 0.216 (0.006) & 0.789 (0.004) & 14.727 & 1.274\\
\hspace{1em}aorsf-fast & 0.216 (0.007) & 0.792 (0.004) & 7.268 & 2.936\\
\hspace{1em}aorsf-cph & 0.215 (0.006) & 0.792 (0.004) & 22.420 & 3.070\\
\hspace{1em}obliqueRSF-net & 0.207 (0.005) & 0.791 (0.004) & 7329.229 & 331.164\\
\hspace{1em}cif-standard & 0.201 (0.004) & 0.790 (0.004) & 72.253 & 384.744\\
\hspace{1em}nn-cox & 0.193 (0.011) & 0.780 (0.005) & 103.296 & 92.992\\
\hspace{1em}glmnet-cox & 0.191 (0.015) & 0.777 (0.007) & 2.256 & 0.011\\
\hspace{1em}ranger-extratrees & 0.181 (0.004) & 0.780 (0.005) & 332.296 & 70.372\\
\hspace{1em}cif-rotate & 0.151 (0.007) & 0.757 (0.006) & 589.958 & 66.749\\
\hspace{1em}xgboost-cox & 0.131 (0.012) & 0.794 (0.004) & 11.220 & 0.015\\
\hspace{1em}aorsf-random & 0.130 (0.005) & 0.733 (0.005) & 19.416 & 2.051\\
\hspace{1em}cif-extension & 0.113 (0.002) & 0.775 (0.005) & 180.483 & 54.092\\
\hspace{1em}xgboost-aft & --- & 0.794 (0.004) & 28.692 & 0.014\\

@@

\hspace{1em}aorsf-fast & 0.233 (0.006) & 0.841 (0.005) & 5.731 & 1.497\\
\hspace{1em}rsf-standard & 0.229 (0.006) & 0.835 (0.005) & 11.238 & 1.079\\
\hspace{1em}aorsf-cph & 0.229 (0.006) & 0.841 (0.005) & 16.768 & 1.549\\
\hspace{1em}aorsf-net & 0.228 (0.006) & 0.841 (0.005) & 597.572 & 1.659\\
\hspace{1em}obliqueRSF-net & 0.212 (0.005) & 0.841 (0.005) & 3737.617 & 320.616\\
\hspace{1em}cif-standard & 0.199 (0.005) & 0.839 (0.005) & 71.854 & 391.419\\

>>>>

This Section introduces negation VI, which is similar to permutation VI in that it measures how much a model’s prediction error increases when a variable’s role in the model is de-stabilized. Specifically, negation VI measures the increase in an oblique RF's prediction error after flipping the sign of all coefficients linked to a variable (\ie, negating them). As negating a coefficient effectively flips decision boundaries around the corresponding predictor's axis, scaling numeric predictors to have a mean of zero and standard deviation of one is recommended.\footnote{The \texttt{aorsf} package automatically scales numeric inputs to a mean of zero and standard deviation of one.} For the current study, we use Harrell's concordance (C)-statistic \citep{harrell1982evaluating} to measure change in prediction error when computing negation VI.

A motivating problem for negation VI is that of correlated, numeric predictors. When two predictors are correlated and only one of them is relevant for prediction of an outcome, VI methods based on permutation are known to have upward bias in estimating the importance of the irrelevant predictor \citep{hooker2021unrestricted}. Permutation may distort the joint distribution of predictors, which may in turn increase a model's estimated prediction error due to extrapolation. On the other hand, negating coefficients modifies the decision rule rather than modifying the data. Thus, we hypothesize that negation VI can be more effective at identifying important predictors for correlated, numeric predictors.

Negation VI has several helpful characteristics. First, negation VI generalizes to any oblique RF (\ie, not just RSFs) using any valid error function, making it both general and flexible.\footnote{The \texttt{aorsf} package enables customized functions to be applied in lieu of the default C-statistic.} Second, negation is non-random and hence reproducible without setting a random seed. Third, since negation VI does not permute variables, the analyst need not worry about impossible combinations of predictors that may occur when one predictor is randomly permuted, such as having a negative status for type 2 diabetes and having Hemoglobin A1c level $\geq$ 6.5\% (a value indicative of type 2 diabetes) as a result of randomly permuting the values of Hemoglobin A1c.

@@

Sections \ref{sec:bm_pred} and \ref{sec:bm_vi} present numerical experiments examining the accelerated oblique RSF and negation VI, respectively. The code used to run these experiments is available online at \href{https://github.com/bcjaeger/aorsf-bench}{https://github.com/bcjaeger/aorsf-bench}. All analyses were conducted using R version 4.1.3 with version 0.0.4 of \texttt{aorsf} \citep{jaeger2022aorsf}, and coordinated by the \texttt{targets} R package \citep{targets}. To standardize comparisons of computational efficiency, all learners and VI techniques used up to 4 processing units.

\subsection{Benchmark of prediction accuracy} \label{sec:bm_pred}

The aim of this numeric experiment is to evaluate the prediction accuracy of the accelerated oblique RSF compared to its predecessor (the oblique RSF from the \texttt{obliqueRSF} R package) and to several other machine learning algorithms. Inferences drawn from this experiment include equivalence and inferiority tests based on Bayesian linear mixed models.

@@

We used a collection of 21 data sets containing a total of 35 risk prediction tasks (tasks per data set ranged from one to four). Participant-level data from the GUIDE-IT and SPRINT clinical trials and the ARIC, MESA, and JHS community cohort studies was obtained from the National Institute of Health Biologic Specimen and Data Repository Coordinating Center (BioLINCC). Designs and protocols for these studies have been made available \citep{aric1989atherosclerosis, bild2002multi, felker2017effect, sprint2015randomized, taylor2005toward}. All other datasets were publicly available and obtained through R packages (see Appendix A.1). Across all prediction tasks, the number of observations ranged from 137 to 17,549 (median: 1,384), the number of predictors ranged from 7 to 1,692 (median: 41), and the percentage of censored observations ranged from 5.26 to 97.7 (median: 78.1) (\tabrefAppendix{tab:datasets}).

@@

For each risk prediction task, we completed 25 runs of Monte-Carlo cross validation. In each run, we used a random sample containing 50\% of the available data for training and the remaining 50\% for testing each of the learners described in \secref{sec:learners}. Then, for each learner, we computed the IPA and time-dependent C-statistic. If any learner failed to obtain predictions on any particular split of data\footnote{For example, when the prediction task was to predict risk of death in the ACTG 320 clinical trial (26 events total), some splits did not leave enough events in the training data to fit complex learners such as neural networks}, the results for that split were omitted from downstream analyses for all learners.

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest IPA in 17 out of 35 risk prediction tasks, with an overall mean IPA of 12.6 (Figure \ref{fig:bm_pred_viz_ibs}). Compared to the learner with the second highest mean IPA (\texttt{cif-standard}), \texttt{aorsf-fast}'s mean was 1.33 points higher, a relative increase of 11.7\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected IPA was 0.99, and the posterior probability of \texttt{aorsf-fast} having a superior IPA to other learners ranged from 0.82 (versus \texttt{cif-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_ibs})

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest time-dependent C-statistic in 9 out of 35 risk prediction tasks, with an overall mean of 77.2 (Figure \ref{fig:bm_pred_viz_cstat}). Compared to the learner with the second highest mean C-statistic (\texttt{cif-standard}), \texttt{aorsf-fast}'s mean was 0.668 points higher, a relative increase of 0.873\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected time-dependent C-statistics was $>$ 0.999, and the posterior probability of \texttt{aorsf-fast} having a superior time-dependent C-statistic versus other learners ranged from 0.14 (versus \texttt{cif-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_cstat})

@@

@@

Parameters that varied in the current simulation study included the number of observations (500, 1000, and 2500) and the absolute value of the maximum correlation between predictors (0.3, 0.15, and 0). Parameters that remain fixed throughout the study included the number of predictors in each class (15) and the effect size of each predictor (one standard deviation increase associated with a 64\% increase in relative risk). Using this design for simulated data, the Heller explained relative risk (95\% confidence interval) of our covariates was 88.5 (88.2, 88.7) \citep{heller2012measure} with 2,500 observations.

@@

\subsection{Benchmark of computational efficiency} \label{sec:bm_compute}

The aim of this numeric experiment is to evaluate the computational efficiency of the accelerated oblique RSF compared to its predecessor (the oblique RSF from the \texttt{obliqueRSF} R package) and to several other machine learning algorithms.

\subsubsection{Evaluation of computational efficiency}

For each learner discussed in \secref{sec:learners} and for each of the 35 risk prediction tasks analyzed in \secref{sec:bm_pred}, we tracked the amount of time required to fit a prediction model (including time used to tune parameters) and compute predicted risk.

We performed additional benchmarks on the time required to fit 500 trees using \texttt{aorsf}, \texttt{randomForestSRC}, and \texttt{ranger}. The learners that represented these R packages were \texttt{aorsf-fast}, \texttt{rsf-standard}, and \texttt{rsf-extratrees}, respectively. To allow for controlled comparisons of computational efficiency with varying dimensions of training data, we used the same process to simulate data as described in \secref{sec:data_sim}, varying the number of observations from 100 to 10000 and the number of predictors from 10 to 1000. The minimum node size of trees in this experiment was dynamically set as the nearest integer to the number of observations in the training set divided by 10.

\subsubsection{Results}

In the analysis of 35 risk prediction tasks, \texttt{aorsf-fast} was the second fastest learner overall, with a median time to develop a risk prediction model and compute predictions about 337 milliseconds longer than \texttt{glmnet-cox} (Figure \ref{fig:bm_pred_time}). Comparing median computing times, \texttt{aorsf-fast} was 1,291.6 times faster than its predecessor, \texttt{obliqueRSF-net}. In addition, \texttt{aorsf-fast} was 19.8, 1.90, and 3.18 faster than axis based forests grown using the \texttt{party}, \texttt{ranger}, and \texttt{randomForestSRC} packages, respectively.

In the analysis of time to fit 500 trees using simulated data, the \texttt{ranger} package exhibited the fastest computation times overall (Figure \ref{fig:bm_time_viz}). \texttt{aorsf} was the second fastest when the number of predictors was 10 or 100,  and \texttt{randomForestSRC} had similar computation time versus \texttt{aorsf} when 1000 predictors were present.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/bm_time_viz-1} \caption[The expected time, in seconds, to fit an ensemble of 500 axis-based survival trees using the ranger or randomForestSRC package versus 500 oblique survival trees using the aorsf package]{The expected time, in seconds, to fit an ensemble of 500 axis-based survival trees using the ranger or randomForestSRC package versus 500 oblique survival trees using the aorsf package. The ranger package is the most efficient overall, and aorsf appears to be relatively efficient in larger samples, particularly when 10 or 100 predictors are present in the training data. All three packages appear to scale linearly in computation time with the number of observations in the training data.}\label{fig:bm_time_viz}
\end{figure}

@@

In this paper, we have developed two contributions to the oblique RSF: (1) the accelerated oblique RSF (\ie, \texttt{aorsf-fast}) and (2) negation VI. Our technique to accelerate the oblique RSF reduces the number of operations required to find linear combinations of inputs using a single iteration of Newton Raphson scoring, while our VI technique directly engages with coefficients in linear combinations of inputs to measure importance of individual variables. In numeric experiments, we found that that \texttt{aorsf-fast} is approximately 1,291.6 times faster than its predecessor, \texttt{obliqueRSF-net}, with a practically equivalent C-statistic. We also found that negation VI, a technique to estimate VI using the oblique RSF, detected non-linear, combination, and main effects more effectively than three standard methods to estimate VI: permutation, ANOVA, and SHAP VI. Overall, we found that estimating VI using negation instead of ANOVA increased the C-statistic for ranking a relevant variable higher than an irrelevant variable by 2.05, a relative increase of 2.78\%.

To understand potential differences in computational efficiency, we reviewed code in the \texttt{aorsf}, \texttt{randomForestSRC}, and \texttt{ranger} packages. We found differences in how survival outcome data are saved in leaf nodes. For each leaf node, \texttt{aorsf} stores data with one row per unique event time using training data that are stored in the leaf, whereas \texttt{randomForestSRC} and \texttt{ranger} store survival outcomes at a fixed grid of event times in each leaf. By default, \texttt{ranger} creates a grid that includes all event times in the training data. The grid strategy can cause higher computing time and memory usage when the grid of event times is large and a large number of leaf nodes are included in each tree, which can occur when minimum node size is small relative to the size of the training data. We kept minimum node size fixed in our benchmark of computational efficiency using real data, and dymanically increased minimum node size based on the size of the training set when we benchmarked computational efficiency using simulated data. Because of this decision, the \texttt{randomForestSRC} and \texttt{ranger} packages ran slower than \texttt{aorsf} in our benchmark of real data but not in the benchmark of simulated data.

@@

\hspace{1em}aorsf-fast & 0.126 (0.109) & 0.772 (0.071) & 0.527 & 0.138\\
\hspace{1em}aorsf-cph & 0.126 (0.109) & 0.771 (0.070) & 1.750 & 0.140\\
\hspace{1em}cif-standard & 0.113 (0.098) & 0.765 (0.071) & 3.792 & 6.448\\
\hspace{1em}rsf-standard & 0.113 (0.114) & 0.755 (0.075) & 1.945 & 0.196\\
\hspace{1em}cif-rotate & 0.112 (0.124) & 0.755 (0.081) & 36.928 & 8.194\\
\hspace{1em}obliqueRSF-net & 0.109 (0.081) & 0.773 (0.069) & 552.335 & 84.685\\
\hspace{1em}glmnet-cox & 0.108 (0.119) & 0.757 (0.077) & 0.326 & 0.003\\
\hspace{1em}ranger-extratrees & 0.099 (0.085) & 0.762 (0.067) & 0.790 & 1.109\\
\hspace{1em}cif-extension & 0.095 (0.092) & 0.761 (0.072) & 21.977 & 6.922\\
\hspace{1em}aorsf-random & 0.092 (0.079) & 0.744 (0.065) & 1.717 & 0.149\\
\hspace{1em}xgboost-cox & 0.064 (0.100) & 0.749 (0.094) & 4.394 & 0.004\\
\hspace{1em}nn-cox & 0.048 (0.106) & 0.657 (0.136) & 11.377 & 1.705\\
\hspace{1em}xgboost-aft & --- & 0.762 (0.076) & 15.031 & 0.007\\

@@

\hspace{1em}obliqueRSF-net & 0.029 (0.015) & 0.753 (0.037) & 115.320 & 18.196\\
\hspace{1em}ranger-extratrees & 0.028 (0.017) & 0.740 (0.036) & 0.086 & 0.133\\
\hspace{1em}aorsf-random & 0.027 (0.020) & 0.756 (0.036) & 0.465 & 0.035\\
\hspace{1em}cif-standard & 0.024 (0.031) & 0.744 (0.040) & 1.657 & 4.377\\
\hspace{1em}aorsf-cph & 0.024 (0.029) & 0.750 (0.042) & 0.436 & 0.036\\
\hspace{1em}aorsf-fast & 0.024 (0.028) & 0.745 (0.045) & 0.141 & 0.036\\
\hspace{1em}cif-extension & 0.023 (0.015) & 0.722 (0.038) & 9.010 & 4.189\\
\hspace{1em}glmnet-cox & 0.016 (0.030) & 0.746 (0.037) & 0.197 & 0.002\\
\hspace{1em}rsf-standard & 0.005 (0.041) & 0.730 (0.042) & 0.179 & 0.061\\
\hspace{1em}cif-rotate & 0.004 (0.040) & 0.731 (0.038) & 14.549 & 3.604\\
\hspace{1em}nn-cox & 0.000 (0.011) & 0.564 (0.101) & 7.755 & 0.811\\
\hspace{1em}xgboost-cox & -0.001 (0.052) & 0.751 (0.033) & 3.729 & 0.003\\
\hspace{1em}xgboost-aft & --- & 0.737 (0.035) & 11.383 & 0.006\\

@@

\hspace{1em}aorsf-random & 0.008 (0.012) & 0.798 (0.073) & 0.285 & 0.024\\
\hspace{1em}obliqueRSF-net & 0.006 (0.009) & 0.821 (0.049) & 49.070 & 11.201\\
\hspace{1em}aorsf-fast & 0.006 (0.019) & 0.826 (0.057) & 0.088 & 0.020\\
\hspace{1em}aorsf-cph & 0.006 (0.018) & 0.818 (0.062) & 0.357 & 0.020\\
\hspace{1em}cif-extension & 0.001 (0.020) & 0.765 (0.066) & 8.283 & 3.478\\
\hspace{1em}ranger-extratrees & 0.001 (0.019) & 0.777 (0.069) & 0.041 & 0.122\\
\hspace{1em}xgboost-cox & -0.004 (0.004) & 0.500 (0.000) & 0.118 & 0.002\\
\hspace{1em}nn-cox & -0.004 (0.004) & 0.547 (0.128) & 7.487 & 0.717\\
\hspace{1em}cif-standard & -0.005 (0.025) & 0.781 (0.062) & 1.695 & 4.223\\
\hspace{1em}rsf-standard & -0.031 (0.051) & 0.776 (0.073) & 0.098 & 0.037\\
\hspace{1em}cif-rotate & -0.037 (0.049) & 0.707 (0.090) & 13.163 & 3.201\\
\hspace{1em}glmnet-cox & -0.065 (0.095) & 0.746 (0.098) & 0.286 & 0.002\\
\hspace{1em}xgboost-aft & --- & 0.774 (0.070) & 10.124 & 0.005\\

@@

\hspace{1em}aorsf-fast & 0.156 (0.007) & 0.810 (0.007) & 4.590 & 1.434\\
\hspace{1em}aorsf-cph & 0.153 (0.006) & 0.809 (0.007) & 14.582 & 1.450\\
\hspace{1em}rsf-standard & 0.150 (0.007) & 0.801 (0.007) & 8.222 & 0.981\\
\hspace{1em}obliqueRSF-net & 0.133 (0.005) & 0.811 (0.008) & 4468.696 & 1359.275\\
\hspace{1em}cif-standard & 0.132 (0.005) & 0.809 (0.007) & 70.462 & 358.273\\
\hspace{1em}glmnet-cox & 0.129 (0.011) & 0.795 (0.008) & 1.873 & 0.010\\
\hspace{1em}nn-cox & 0.126 (0.012) & 0.795 (0.007) & 43.144 & 86.104\\
\hspace{1em}ranger-extratrees & 0.112 (0.005) & 0.795 (0.009) & 283.364 & 69.939\\
\hspace{1em}aorsf-random & 0.104 (0.005) & 0.771 (0.008) & 11.343 & 1.420\\
\hspace{1em}cif-rotate & 0.104 (0.004) & 0.783 (0.009) & 558.249 & 68.510\\
\hspace{1em}cif-extension & 0.069 (0.002) & 0.786 (0.009) & 164.161 & 50.420\\
\hspace{1em}xgboost-cox & 0.064 (0.017) & 0.813 (0.006) & 9.660 & 0.015\\
\hspace{1em}xgboost-aft & --- & 0.814 (0.006) & 29.321 & 0.013\\

@@

\hspace{1em}rsf-standard & 0.216 (0.006) & 0.789 (0.004) & 12.352 & 1.162\\
\hspace{1em}aorsf-fast & 0.216 (0.006) & 0.792 (0.004) & 7.463 & 2.519\\
\hspace{1em}aorsf-cph & 0.215 (0.006) & 0.792 (0.004) & 22.389 & 2.569\\
\hspace{1em}cif-standard & 0.201 (0.004) & 0.790 (0.004) & 68.083 & 373.333\\
\hspace{1em}obliqueRSF-net & 0.195 (0.004) & 0.789 (0.004) & 7373.936 & 1276.463\\
\hspace{1em}nn-cox & 0.191 (0.008) & 0.779 (0.005) & 83.418 & 84.770\\
\hspace{1em}glmnet-cox & 0.191 (0.015) & 0.777 (0.007) & 2.282 & 0.011\\
\hspace{1em}ranger-extratrees & 0.181 (0.004) & 0.780 (0.005) & 356.276 & 61.525\\
\hspace{1em}cif-rotate & 0.151 (0.007) & 0.757 (0.006) & 563.575 & 64.849\\
\hspace{1em}xgboost-cox & 0.131 (0.012) & 0.794 (0.004) & 12.568 & 0.015\\
\hspace{1em}aorsf-random & 0.128 (0.006) & 0.725 (0.005) & 20.291 & 2.298\\
\hspace{1em}cif-extension & 0.113 (0.002) & 0.775 (0.005) & 176.867 & 50.345\\
\hspace{1em}xgboost-aft & --- & 0.794 (0.004) & 35.380 & 0.013\\

@@

\hspace{1em}aorsf-fast & 0.233 (0.006) & 0.841 (0.005) & 5.478 & 1.709\\
\hspace{1em}rsf-standard & 0.229 (0.006) & 0.835 (0.005) & 10.326 & 1.039\\
\hspace{1em}aorsf-cph & 0.229 (0.006) & 0.840 (0.005) & 17.063 & 1.730\\
\hspace{1em}cif-standard & 0.199 (0.005) & 0.839 (0.005) & 68.929 & 364.202\\
\hspace{1em}obliqueRSF-net & 0.198 (0.005) & 0.839 (0.005) & 5341.195 & 1624.560\\
\hspace{1em}nn-cox & 0.188 (0.018) & 0.826 (0.006) & 58.351 & 89.012\\
\hspace{1em}cif-rotate & 0.172 (0.006) & 0.806 (0.007) & 569.166 & 67.654\\
